\documentclass[12pt,a4paper,oneside]{report}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{tocloft}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{float}
\usepackage{xcolor}
\usepackage{listings}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{appendix}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{lipsum}

% ── Page geometry ─────────────────────────────────────────────────────────────
\geometry{margin=1in}
\onehalfspacing

% ── Header / Footer ───────────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[R]{\thepage}
\fancyhead[L]{\leftmark}
\renewcommand{\headrulewidth}{0.4pt}

% ── Chapter formatting ────────────────────────────────────────────────────────
\titleformat{\chapter}[display]
  {\normalfont\bfseries\Large\centering}
  {\chaptertitlename\ \thechapter}{12pt}{}
\titlespacing*{\chapter}{0pt}{-20pt}{20pt}

% ── Code listings ─────────────────────────────────────────────────────────────
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny,
  xleftmargin=2em,
  language=Python,
  captionpos=b,
}

% ── Custom commands ───────────────────────────────────────────────────────────
\newcommand{\term}[1]{\textit{#1}}

% ── Document info ─────────────────────────────────────────────────────────────
\title{Integrated agricultural product data System}
\author{%
  Kidan Dibekulu \quad Kaleab Kebede \\
  Kalkidan Yishak \quad Dagmawi Shewaye
}

\begin{document}

% ==============================================================================
% TITLE PAGE
% ==============================================================================
\begin{titlepage}
\begin{center}

\vspace*{2cm}

{\LARGE\bfseries ADDIS ABABA SCIENCE AND TECHNOLOGY UNIVERSITY}\\[0.5cm]
{\large College of Engineering}\\[0.2cm]
{\large Department of Electrical and Computer Engineering}\\[1.5cm]

{\Huge\bfseries Integrated agricultural product data System}\\[1.5cm]

{\large Final Year Project Report}\\[1cm]

\begin{tabular}{ll}
\textbf{Name} & \textbf{ID} \\
Kidan Dibekulu  & ETS0916/14 \\
Kaleab Kebede   & ETS0857/14 \\
Kalkidan Yishak & ETS0888/14 \\
Dagmawi Shewaye & ETS0425/14 \\
\end{tabular}\\[1.5cm]

{\large Submitted to Addis Ababa Science and Technology University}\\
{\large for the award of Bachelor of Science Degree in}\\
{\large Electrical and Computer Engineering (Computer Engineering)}\\[1cm]

{\large Advisor: Ir. Esubalew M.}\\[1cm]

{\large June 2026}

\end{center}
\end{titlepage}

% ==============================================================================
% DECLARATION
% ==============================================================================
\chapter*{Declaration}
\addcontentsline{toc}{chapter}{Declaration}

We, the undersigned, hereby declare that the project work entitled ``Integrated agricultural product data System'' is our original work and has been carried out in partial fulfillment of the requirements for the Bachelor of Science Degree in Computer Engineering at Addis Ababa Science and Technology University.

This project has not been submitted, in whole or in part, to any other institution or university for the award of any academic degree or diploma. All sources of information used during the course of this research have been duly acknowledged and referenced.

The results, analyses, and conclusions presented in this report are based on our own implementation, experimentation, and findings, conducted under the supervision of Ir. Esubalew M.

We understand that any violation of academic honesty and integrity may lead to disciplinary action as per the university's regulations.

\vspace{1cm}
\begin{tabular}{lll}
\textbf{Name} & \textbf{ID Number} & \textbf{Signature} \\[0.3cm]
1. Dagmawi Shewaye  & ETS0425/14 & \rule{5cm}{0.4pt} \\[0.3cm]
2. Kaleab Kebede    & ETS0857/14 & \rule{5cm}{0.4pt} \\[0.3cm]
3. Kalkidan Yishak  & ETS0888/14 & \rule{5cm}{0.4pt} \\[0.3cm]
4. Kidan Dibekulu   & ETS0916/14 & \rule{5cm}{0.4pt} \\
\end{tabular}

% ==============================================================================
% ACKNOWLEDGEMENT
% ==============================================================================
\chapter*{Acknowledgement}
\addcontentsline{toc}{chapter}{Acknowledgement}

First and foremost, we would like to express our deepest gratitude to the Almighty God for granting us the strength, perseverance, and wisdom to complete this project successfully.

We are sincerely thankful to our advisor, Mr.\ Esubalew M., for his continuous guidance, insightful feedback, and dedicated support throughout the course of this project. His expertise in the field of software engineering and system architecture greatly shaped the direction and quality of our work.

We would also like to thank the faculty and staff of the Department of Electrical and Computer Engineering, Addis Ababa Science and Technology University, for providing us with the academic resources, laboratory infrastructure, and a supportive learning environment that enabled us to conduct this research.

Our heartfelt appreciation also goes to the creators and maintainers of open-source technologies including AdonisJS, Next.js, FastAPI, scikit-learn, Leaflet.js, Recharts, and PostgreSQL, which played a vital role in building and validating our system.

We are immensely grateful to our families, friends, and peers for their unwavering encouragement, motivation, and moral support during challenging times. Their belief in us has been a constant source of inspiration.

Finally, this work is the result of collective effort, dedication, and shared vision, and we are proud to present it as a contribution to the advancement of digital agriculture and intelligent systems in Ethiopia.

% ==============================================================================
% TABLE OF CONTENTS
% ==============================================================================
\tableofcontents
\newpage

% ==============================================================================
% LIST OF FIGURES
% ==============================================================================
\listoffigures
\newpage

% ==============================================================================
% LIST OF TABLES
% ==============================================================================
\listoftables
\newpage

% ==============================================================================
% ABBREVIATIONS
% ==============================================================================
\chapter*{List of Abbreviations}
\addcontentsline{toc}{chapter}{List of Abbreviations}

\begin{tabular}{ll}
AI     & Artificial Intelligence \\
API    & Application Programming Interface \\
CNN    & Convolutional Neural Network \\
CRUD   & Create, Read, Update, Delete \\
CSV    & Comma-Separated Values \\
DB     & Database \\
FAO    & Food and Agriculture Organization \\
GB     & Gradient Boosting \\
HTTP   & Hypertext Transfer Protocol \\
IADS   & Integrated agricultural product data System \\
JWT    & JSON Web Token \\
LLM    & Large Language Model \\
MAE    & Mean Absolute Error \\
ML     & Machine Learning \\
MVC    & Model-View-Controller \\
NGO    & Non-Governmental Organization \\
ORM    & Object-Relational Mapping \\
RAG    & Retrieval-Augmented Generation \\
RBAC   & Role-Based Access Control \\
RMSE   & Root Mean Square Error \\
REST   & Representational State Transfer \\
SQL    & Structured Query Language \\
UI     & User Interface \\
UUID   & Universally Unique Identifier \\
VPS    & Virtual Private Server \\
XLSX   & Microsoft Excel Open XML Spreadsheet \\
\end{tabular}

\newpage

% ==============================================================================
% ABSTRACT
% ==============================================================================
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

agricultural product data collection and management in Ethiopia face significant challenges due to unreliable internet connectivity, fragmented record-keeping systems, and lack of access to predictive analytics. Traditional methods relying on paper-based surveys are time-consuming, error-prone, and inaccessible in many rural and under-resourced areas. This study presents the development of an Integrated agricultural product data System, an offline-capable geospatial crop monitoring platform that combines offline-first architecture, AI-powered analytics, and stakeholder-specific dashboards to address these challenges.

The system integrates five core components: an offline data synchronization module using a queue-and-check mechanism for reliable field data collection, a legacy data import tool for bulk ingestion of CSV and Excel files with automatic schema detection, a scikit-learn-based Gradient Boosting Regressor for yield prediction achieving a Mean Absolute Error of 863.54~kg/ha and RMSE of 1539.96~kg/ha, a Retrieval-Augmented Generation system using Anthropic Claude or Cerebras LLMs enabling natural language queries on agricultural product data sets, and stakeholder-specific dynamic dashboards tailored for government, NGO, trader, and researcher users with role-based access control.

The training dataset consisted of 2,000 crop yield records spanning 2015 to 2024 across six major Ethiopian regions (Afar, Amhara, Oromia, Sidama, SNNPR, Tigray) covering eight crop types (teff, wheat, maize, barley, sorghum, sesame, chickpea, coffee) across two growing seasons (Meher and Belg). The Gradient Boosting model was trained on seven features including area, rainfall, temperature, fertilizer amount, year, crop type, and season.

Integrated into a full-stack web application built with AdonisJS and Next.js, the system enables field agents to collect data with offline synchronization, upload historical datasets, and allows stakeholders to query agricultural product data using natural language through the RAG interface. The system implements role-based dashboards with interactive maps using Leaflet.js and charts using Recharts, along with comprehensive audit logging and data governance features.

This project contributes to the field of digital agriculture by offering a comprehensive, accessible, and accurate platform that empowers stakeholders to make timely decisions, reduce crop losses, and promote sustainable farming practices in connectivity-constrained environments, directly supporting Ethiopia's Digital Ethiopia 2025 strategy.

% ==============================================================================
% CHAPTER 1: INTRODUCTION
% ==============================================================================
\include{chapters/01-introduction}

% ==============================================================================
% CHAPTER 2: LITERATURE REVIEW
% ==============================================================================
\include{chapters/02-literature-review}

% ==============================================================================
% CHAPTER 3: METHODOLOGY AND DESIGN
% ==============================================================================
\include{chapters/03-methodology}

% ==============================================================================
% CHAPTER 4: RESULT AND DISCUSSION
% ==============================================================================
\include{chapters/04-results}

% ==============================================================================
% CHAPTER 5: CONCLUSION AND RECOMMENDATIONS
% ==============================================================================
\include{chapters/05-conclusion}

% ==============================================================================
% BIBLIOGRAPHY
% ==============================================================================
\bibliographystyle{ieeetr}
\bibliography{references}

% ==============================================================================
% APPENDICES
% ==============================================================================
\begin{appendix}
\include{chapters/A-appendix}
\end{appendix}

\end{document}
